### 2016-03-31 Manuscript/BiologyLetters/Missing_mammals-draft.tex
Revision 2 round 2

The issues above highlight that it is crucial to have sufficient coded anatomical characters available for living taxa in a clade before using the Total Evidence approach.
However, it is unclear how many coded anatomical characters are actually available for living taxa, i.e. already coded from museum specimens and deposited in phylogenetic matrices accessible online, and how this data is distributed across clades.
Intuitively, most people assume this data has already been collected, but empirical analyses suggest otherwise (e.g. in \cite{ronquista2012,slaterphylogenetic2013,beckancient2014}).
To investigate this further, we assess the number of available coded anatomical characters for living mammals to determine whether enough data exists to build reliable Total Evidence phylogenies.
We also determine whether the characters are phylogenetically overdispersed or clustered across mammalian orders.
%TG: a point to be made here maybe is that we do NOT look at hard tissue characters, rather just the presence of any characters (which is already poor enough)?
%TG: on the other hand, we might not need to emphasize that to much since, anyway, it is clearly explained in the methods (i.e. we tool all cladistic data from matrices)
% NC: If someone makes a point about it in a review it means it isn't explained clearly enough. I think adding a note to the methods and/or discussion is needed.
%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------
% NC: From here on in we need to replace the word cladistic. I don't like the "anatomical characters" as it's a bit long. How about character matrices? You decide.
% TO DO
\section{Materials and Methods}
\subsection{Data collection and standardisation}
We downloaded all discrete morphological matrices containing any living and/or fossil mammal taxa from three major public databases: MorphoBank (\url{morphobank.org} \cite{morphobank}), Graeme Lloyd's website (\url{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\url{github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search for matrices that were not uploaded to these databases (see Electronic Supplementary Material (ESM) for details).
In total, we downloaded 286 matrices containing 5228 unique operational taxonomic units (OTUs). 
We used OTUs rather than species because entries in the matrices ranged from species to families. 
We standardised the taxonomy as described in the ESM and excluded OTUs that were not present in the phylogeny of \cite{BinindaEmonds} or the taxonomy of \cite{wilson2005mammal} to remove fossil species.
This resulted in 1601 unique OTUs from 286 matrices.

@@

% NC: The ref for our paper looks weird - it has a url, probably fix that and check the reference is correct so we get the citation.
% Ref 10 - there's an 03 after the date. Not sure what that refers to.
% NC: Ref 16 - is KPg mising a - ?
% NC: Ref 17 has a missing space

>>>>

The issues above highlight that it is crucial to have sufficient coded anatomical characters available for living taxa in a clade before using the Total Evidence approach.
However, it is unclear how many coded anatomical characters are actually available for living taxa, i.e. already coded from museum specimens and deposited in phylogenetic matrices accessible online, and how this data is distributed across clades.
Intuitively, most people assume this data has already been collected, but empirical analyses suggest otherwise (e.g. in \cite{ronquista2012,slaterphylogenetic2013,beckancient2014}).
To investigate this further, we assess the number of available coded anatomical characters for living mammals to determine whether enough data exists to build reliable Total Evidence phylogenies.
We also determine whether the characters are phylogenetically overdispersed or clustered across mammalian orders.
%TG: a point to be made here maybe is that we do NOT look at hard tissue characters, rather just the presence of any characters (which is already poor enough)?
%TG: on the other hand, we might not need to emphasize that to much since, anyway, it is clearly explained in the methods (i.e. we tool all cladistic data from matrices)
% NC: If someone makes a point about it in a review it means it isn't explained clearly enough. I think adding a note to the methods and/or discussion is needed.
%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------
% NC: From here on in we need to replace the word cladistic. I don't like the "anatomical characters" as it's a bit long. How about character matrices? You decide.
% TO DO
\section{Materials and Methods}
\subsection{Data collection and standardisation}
We downloaded all discrete morphological matrices containing any living and/or fossil mammal taxa from three major public databases: MorphoBank (\url{morphobank.org} \cite{morphobank}), Graeme Lloyd's website (\url{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\url{github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search for matrices that were not uploaded to these databases (see Electronic Supplementary Material 1 (ESM1) for details).
In total, we downloaded 286 matrices containing 5228 unique operational taxonomic units (OTUs). 
We used OTUs rather than species because entries in the matrices ranged from species to families. 
We standardised the taxonomy as described in the ESM1 and excluded OTUs that were not present in the phylogeny of \cite{BinindaEmonds} or the taxonomy of \cite{wilson2005mammal} to remove fossil species.
This resulted in 1601 unique OTUs from 286 matrices.

@@

% NC: Ref 16 - is KPg mising a - ?

### 2016-03-31 Manuscript/BiologyLetters/response2.tex
Revision 2 round 2

We emphasised the importance of skeletal characters in the introduction as follows:
``A downside of this method is that it requires molecular data for living taxa and discrete morphological/anatomical data shared between both living and fossil taxa (i.e. hard tissue characters such as skeletal features).''
lines @@@.

We also added a comment in the discussion to the effect that having data for a living mammal doesn't necessarily mean it can overlap with a fossil - it could all be soft tissue characters. However, in general our datasets appear to contain at least some hard character for each living species so it shouldn't influence our results. But yes it is an important consideration for future work and we thank the reviewer for pointing this out (we misunderstood this point in the first round of revisions).

% NC: Need to add a line or two for this in the discussion. Or maybe rephrase the above and we can just stick with the clarifications I've added to the methods and intro.
% TO DO

% NC: I've changed a few bits in the text too
% TO DO

@@

We agree with this comment (and Referee 3 comment 2) and modified the paragraph discussing this caveat as follows:
``Although this will bias our results towards lower coverage we do not think this bias will be large, as many recent morphological matrices reuse living taxa characters from older matrices (see ESM1), so many of these data will be present in our analyses.
Additionally, these older matrices are likely to differ from more recent ones in terms of their underlying definition of homology and their coding practices (see [19]).
Therefore, care needs to be taken when deciding how to include these older matrices.
''

lines @@@.

% NC: I've editted this a bit to get more closely at what the reviewer wants (it thinks OUR analysis might be biased, and is correct).
% Make sure you also write a brief explanation here as well as editting the text. Perhaps add a line like:
% TO DO
We are unable to properly quantify the bias introduced by ignoring older matrices without going out and collecting them all exhaustively and re-running our analyses. This would take a substantial amount of time not available at present.

\item{\textcolor{blue}{I don't agree with the assertion that the genus level should be our primary concern; more and more studies are building species-level trees with fossil taxa. I'm happy for the authors to discuss the patterns at the genus level, because it is one of the important levels, but given that genera are an artificial line in the taxonomic sand, it's certainly not the primary one, and I'd hate to see this statement sail unopposed into the literature. I'd like to see it moderated a bit in the final version of the manuscript.}}
Following Referee's 1 and 4's comments, we removed this paragraph.

@@

\begin{enumerate}
\item{\textcolor{blue}{line 34: spelling: ccombines = combines}}
We fixed this typo.

\item{\textcolor{blue}{lines 44 - 52. I'd suggest rewording this a little bit as lines 48 - 52 are a little awkward. May I suggest the following:
"We might expect the total evidence method to perform poorly when there is little overlap between coded anatomical characters in living and fossil taxa. This is because fossil taxa cannot be correctly placed within a clade of living species if none of its members have been coded for morphological characters. Furthermore, simulations show that fossils are more likely to be placed in clades for which more characters have been coded, regardless of whether this is the correct clade [8]."}}
We replaced the section with Referee 2's suggestion.

\item{\textcolor{blue}{line 152: ... a carnivoran fossil is unlikely to be placed in Herpestidae because ....}}
We fixed this typo.

\item{\textcolor{blue}{line 159: I'd replace "It is worth noting" with "We acknowledge"}}
We changed the start of the sentence as suggested.

\item{\textcolor{blue}{line 161: word order: provided easily accessible matrices}}
We changed the word order.

@@

\begin{enumerate}
\item{\textcolor{blue}{The new title is an improvement.  However, I think that it could be altered still further: would something like: "Assessment of available anatomical characters for linking living mammals to fossil taxa in phylogenetic analyses"?  That is the main thrust of the paper, after all, and it also avoids having someone simply think: "Who cares? I have molecules" and moving on to the next article.}}
We changed the title as suggested by Referee 3.

\item{\textcolor{blue}{Lines 159 - 160: "It is worth noting, however, that our analysis did not include all the matrices containing anatomical characters ever published."
One thing to note is that a lot of the older analyses, which tended to be morphological, were done in the early days of modern phylogenetics, and those are the matrices that are most apt to not have pdf or html versions.  So, there might be a bias against the "pioneering" studies.  However, many of those studies will be a quarter of a century old, and the homology concepts and (especially!) coding protocols being used back then almost certainly differ from now.  Also, a lot of those matrices were "cleaned" to remove homoplasy: this was not attempted fraud, but due to the fact that computer runs that are done in hours today took weeks and weeks back then, and actually storing all of the trees that might come out of "messy" matrices was beyond the old floppy drives.  That is one reason why I rarely use matrices published before the early 1990's in my own meta-analyses.  So, if those matrices do exist and can be computerized, then I think that they should be viewed as "starting points" for coding taxa: they almost certainly will need updating.}}

% TO DO (see Referee 1 comment 2)
%NC: Also need to add a line sayong we've now mentioned this in the discussion and add the bit here.

@@

\item{\textcolor{blue}{In particular I agree with Referee 1 regarding the continued use of "cladistic" and the genus-level issue (many palaeontologists are now focused on the species-level and certainly many mammalian genera are not monotypic - this is more of an issue with dinosaur workers), and Referee 3 (Pete Wagner's) title suggestion is - I think - more likely to attract readers.}}
We dealt with these suggestions (see respectively Referee 1 comment 1, Referee 1 comment 3 and Referee 3 comment 1).

\item{\textcolor{blue}{on L174 ("but also to any researcher focusing understanding macroevolutionary patterns and processes"), this seems to be missing an "on".}}
We fixed this typo.
\end{enumerate}

>>>>

We emphasised the importance of skeletal characters in the introduction as follows:
\textit{``A downside of this method is that it requires molecular data for living taxa and discrete morphological/anatomical data shared between both living and fossil taxa (i.e. hard tissue characters such as skeletal features).''}
lines @@@.

We also added a comment in the discussion to the effect that having data for living mammals doesn't necessarily mean it can overlap with a fossil - it could all be soft tissue characters for example. However, in general our datasets appear to contain at least some hard character for each living species so it shouldn't influence our results. But yes it is an important consideration for future work and we thank the reviewer for pointing this out (we misunderstood this point in the first round of revisions).

We add the following sentence to the discussion:
\textit{``Additionally, our analyses may underestimate the problem as we do not distiguish between soft and hard tissue characters; if a living taxon has only soft tissue coded anatomical characters then it will not have overlapping data with fossils that only have hard tissues preserved.''}
lines @@@.

@@

We agree with this comment (and Referee 3 comment 2) and modified the paragraph discussing this caveat as follows:
\textit{``Although this will bias our results towards lower coverage we do not think this bias will be large, as many recent morphological matrices reuse living taxa characters from older matrices (see ESM1), so many of these data will be present in our analyses.
Additionally, these older matrices are likely to differ from more recent ones in terms of their underlying definition of homology and their coding practices (see [19]).
Therefore, care needs to be taken when deciding how to include these older matrices.''} lines @@@.

Since we did not collect these ``older'' matrices, we are unable to properly quantify the bias introduced by ignoring them.
To test it we should rerun the whole analysis from the data collection to include them and to see the difference between the present data and the ones including these ``older'' matrices.
This would really interesting but would also take a substantial amount of time not available at present.

\item{\textcolor{blue}{I don't agree with the assertion that the genus level should be our primary concern; more and more studies are building species-level trees with fossil taxa. I'm happy for the authors to discuss the patterns at the genus level, because it is one of the important levels, but given that genera are an artificial line in the taxonomic sand, it's certainly not the primary one, and I'd hate to see this statement sail unopposed into the literature. I'd like to see it moderated a bit in the final version of the manuscript.}}

Following Referee's 1 and 4's comments, we removed this paragraph.

@@

\begin{enumerate}
\item{\textcolor{blue}{line 34: spelling: ccombines = combines}}

We fixed this typo.

\item{\textcolor{blue}{lines 44 - 52. I'd suggest rewording this a little bit as lines 48 - 52 are a little awkward. May I suggest the following:
``We might expect the total evidence method to perform poorly when there is little overlap between coded anatomical characters in living and fossil taxa. This is because fossil taxa cannot be correctly placed within a clade of living species if none of its members have been coded for morphological characters. Furthermore, simulations show that fossils are more likely to be placed in clades for which more characters have been coded, regardless of whether this is the correct clade [8].''}}

We replaced the section with Referee 2's suggestion.

\item{\textcolor{blue}{line 152: ... a carnivoran fossil is unlikely to be placed in Herpestidae because ....}}

We fixed this typo.

\item{\textcolor{blue}{line 159: I'd replace ``It is worth noting'' with``We acknowledge''}}

We changed the start of the sentence as suggested.

\item{\textcolor{blue}{line 161: word order: provided easily accessible matrices}}

We changed the word order.

@@

\begin{enumerate}
\item{\textcolor{blue}{The new title is an improvement.  However, I think that it could be altered still further: would something like: ``Assessment of available anatomical characters for linking living mammals to fossil taxa in phylogenetic analyses``"''?  That is the main thrust of the paper, after all, and it also avoids having someone simply think: ``Who cares? I have molecules'' and moving on to the next article.}}

We changed the title as suggested by Referee 3.

\item{\textcolor{blue}{Lines 159 - 160: ``It is worth noting, however, that our analysis did not include all the matrices containing anatomical characters ever published''.
One thing to note is that a lot of the older analyses, which tended to be morphological, were done in the early days of modern phylogenetics, and those are the matrices that are most apt to not have pdf or html versions.  So, there might be a bias against the ``pioneering``"'' studies.  However, many of those studies will be a quarter of a century old, and the homology concepts and (especially!) coding protocols being used back then almost certainly differ from now.  Also, a lot of those matrices were ``cleaned``"'' to remove homoplasy: this was not attempted fraud, but due to the fact that computer runs that are done in hours today took weeks and weeks back then, and actually storing all of the trees that might come out of ``messy'' matrices was beyond the old floppy drives.  That is one reason why I rarely use matrices published before the early 1990's in my own meta-analyses.  So, if those matrices do exist and can be computerized, then I think that they should be viewed as ``starting points``"'' for coding taxa: they almost certainly will need updating.}}

We dealt with this comment above and mentioned this in the discussion (see Referee 1, comment 2).

@@

\item{\textcolor{blue}{In particular I agree with Referee 1 regarding the continued use of ``cladistic``"'' and the genus-level issue (many palaeontologists are now focused on the species-level and certainly many mammalian genera are not monotypic - this is more of an issue with dinosaur workers), and Referee 3 (Pete Wagner's) title suggestion is - I think - more likely to attract readers.}}

We dealt with these suggestions (see respectively Referee 1 comment 1, Referee 1 comment 3 and Referee 3 comment 1).

\item{\textcolor{blue}{on L174 ("but also to any researcher focusing understanding macroevolutionary patterns and processes"), this seems to be missing an ``on``"''.}}

We fixed this typo.
\end{enumerate}
